Supplementary Information

### 2021-05-05 doc/manuscript/ForestGEO_treerings.Rmd
intro, table formatting and automation thereof

## DBH
Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @enquist_global_2002; @niklas_plant_2004; **REF**] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
Despite the fundamental importance of a solid understanding of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
In addition, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in cross-sectional analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

@@

Rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings. 
Approaches that ...., such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].

@@

\newpage
**Table 1 | (DRAFT hypothesis table-- needs to be filled in and condensed/ cleaned up)**. .... 
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)
hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypotheses_table <- subset(hypotheses_table, select = -c(table_row, Q, H, category, notes) )#remove columns that shouldn't be included
ft <- flextable(hypotheses_table)
ft <- italic(ft, i=c(1,15,22), j=1)
ft <- bold(ft, i=c(1,15,22, 2,6,12,16,19, 23, 25))
ft <- line_spacing(ft, space = 1, part = "all")
ft <- autofit(ft)
ft <- footnote( ft, i = 1, j = 2,
                value = as_paragraph(c("SSC= species-site combinations; SSCV=species-site-variable combination")),
                ref_symbols = c("*"),
                part = "header",inline=T)
ft

>>>>

## DBH
Tree stem diameter, DBH, scales predictably with numerous traits affecting tree physiology [e.g., height, crown size and position, root mass, hydraulic architecture, @enquist_global_2002; @niklas_plant_2004; **REF**] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
To use tree rings to predict tree growth, ecosystem productivity, and forest dynamics, we need models that include DBH [e.g., @evans_fusing_2017; @klesse_amplifying_2020] and its potential climate interactions [e.g., @rollinson_climate_2021].
Despite the fundamental importance of a solid understanding of the relationship between DBH and growth, there remain inconsistencies as to what is considered a typical growth pattern. 
In dendrochonology, the "textbook" pattern of radial stem growth increments (ring width, RW) is one of rapid initial growth followed by a slow decline [@fritts_tree_1976], whereas forest census data reveals a pattern of increasing RW with DBH across globally distributed forests [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
In addition, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
A solid understanding of the ontogeny of tree growth is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

@@

Rising atmospheric CO~2~ concentrations, and changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings. 
Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
Yet, challenges remain...

@@

\newpage
**Table 1 | (DRAFT hypothesis table-- needs to be filled in and condensed/ cleaned up)**. .... 
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(flextable)
set_flextable_defaults(font.family = "Arial", font.size = 9)
hypotheses_table <- read.csv("tables_figures/hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
question_rows<- which(!is.na(hypotheses_table$Q) )
hypothesis_rows<-which(!is.na(hypotheses_table$H) )

hypotheses_table <- subset(hypotheses_table, select = -c(Q, H, category, notes) )#remove columns that shouldn't be included
ft <- flextable(hypotheses_table)
ft <- italic(ft, i=question_rows, j=1)
ft <- bold(ft, i=c(question_rows, hypothesis_rows))
ft <- line_spacing(ft, space = 1, part = "all")
ft <- padding(ft, padding.top = 2)
ft <- padding(ft, padding.bottom = 2)
ft <- padding(ft, j = 1, padding.left = 20)
ft <- padding(ft, i = hypothesis_rows, j = 1, padding.left = 10)
ft <- padding(ft, i = question_rows, j = 1, padding.left = 0)
ft <- autofit(ft)
ft <- footnote( ft, i = 1, j = 2,
                value = as_paragraph(c("SSC= species-site combinations; SSCV=species-site-variable combination")),
                ref_symbols = c("*"),
                part = "header",inline=T)
ft
